
---
output: html_document
---
---
title: "Coursera: Regression Models - Project"
subtitle: "Motor Trend Car Road Tests"
author: "Sumant Sharma"

---  

***

### Executive Summary

In this project we work for Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, we explor the relationship between a set of variables and miles per gallon (MPG) (outcome). We are particularly interested in the following two questions:

* Is an automatic or manual transmission better for <code>MPG</code>?
* Quantify the <code>MPG</code> difference between 'automatic' and 'manual' transmissions.

***
### Solution   
#### 1. Data Dictionary   
This dataset was extracted from the 1974 Motor Trend US magazine which comprises of 11 aspects of automobile design and performance for 32 automobiles. Refer Table 1(Appendix - A) for the dataset.
```{r, results='hide'}
data(mtcars)
head(mtcars)
```   


#### 2. Data Transformation   
The first step to build any data model is to cleanse the data. As part of data cleansing, we transformed and factored the following categorical variables.   

```{r, echo=TRUE}
data(mtcars)
mtcars$cyl <- factor(mtcars$cyl)
mtcars$vs <- factor(mtcars$vs)
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
mtcars$am <- factor(mtcars$am,labels=c('Automatic','Manual'))
```   

#### 3. Data Exploration   
Once data transformation was complete, we used multiple data exploration tools such as scatterplots, histograms etc. to identifying meaningful relationships among variables.   
   
<code>scatterPlotMatrix(mtcars)</code>   
   
From the Plot 1(Appendix - B) we notice that variables like <code>cyl</code>, <code>disp</code>, <code>hp</code>, <code>drat</code>, <code>wt</code>, <code>vs</code> and <code>am</code> seem to have some strong correlation with <code>mpg</code>. We also plot a boxplot of the variable <code>mpg</code> grouped by <code>am</code>. Plot 2(Appendix - B) shows that the mpg is higher when the transmission is 'Manual'.   
   
<code>library(ggplot2)  
p <- ggplot(mtcars, aes(am, mpg))  
p + geom_boxplot(aes(fill=am))  
</code>

#### 4. Data Modeling   
Let's investigate the relationship between miles per gallon and transmission type. First we run a regression model with <code>mpg</code> as the response variable and <code>am</code> as the (only) explanatory variable.   
```{r, results='hide'}
mod1<-lm(mpg ~ am, data = mtcars)
summary(mod1)$coef
```   
Table 2(Appendix - A) shows that the model explains only 35% of the total variation in <code>mpg</code> and there are additional variables in our data; there is a chance that we could find a better model. So we run the regression model multiple times using STEP method to build multiple regression models using both forward selection and backward elimination methods by the AIC algorithm.   
```{r, results='hide'}
mod2<-lm(mpg ~ ., data = mtcars)
mod3<-step(mod2, direction = "both")
summary(mod3)
```   
We can see in Table 3(Appendix - A) the adjusted R^2^ value is equal to 0.84, thus this model can explain 84% of the variability with <code>cyl</code>, <code>wt</code>, <code>hp</code> and <code>am</code> as relevant variables.   
We further compare the models with <code>ANOVA</code>. Based on the significant p-values of model coefficients, the model <code>(mpg ~ cyl + hp + wt + am)</code> is our final selection - refer Table 4(Appendix - A).

```{r, results='hide'}
anova(mod1, mod3)
```

#### 5. Residuals
<code>par(mfrow = c(2,2))
plot(mod3)
</code> 
Plot 3(Appendix - B) depicts that the residual plots are  randomly scattered.   

#### 6. Conclusion
From the <code>mod3 (mpg ~ cyl + hp + wt + am)</code> model we can conclude that 'Manual' transmission is better than 'Automatic' transmission for <code>mpg</code>. Miles per gallon <code>mpg</code> will increase by 1.81 in cars with 'Manual' transmission in comparison to cars with 'Automatic' transmission (adjusted by <code>cyl, hp & wt</code>).   

***
#### Appendix - A
```{r loading, cache=TRUE, echo=FALSE, fig.width=6}
data(mtcars)
head(mtcars)
```   

Table 1: mtcars dataset

```{r, echo=FALSE, fig.width=6}
mod1<-lm(mpg ~ am, data = mtcars)
summary(mod1)
```   

Table 2: Summary of the model with <code>mpg</code> as the response variable with <code>am</code> as predictor.

```{r, echo=FALSE, fig.width=6}
mod2<-lm(mpg ~ cyl + hp + wt + am, data = mtcars)
summary(mod2)
```   

Table 3: Summary of the model with <code>mpg</code> as the response variable by <code>step</code> method.

```{r, echo=FALSE, fig.width=6}
anova(mod1, mod3)
```

Table 4: Analysis of variance between <code>mod1</code> and <code>mod3</code>   

***

#### Appendix - B

```{r, echo=FALSE, fig.height=10, fig.width=10}
panel.cor.scale <- function(x, y, digits=2, prefix="", cex.cor)
{
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r = (cor(x, y,use="pairwise"))
  txt <- format(c(r, 0.123456789), digits=digits)[1]
	txt <- paste(prefix, txt, sep="")
	if(missing(cex.cor)) cex <- 0.5/strwidth(txt)
	text(0.5, 0.5, txt, cex = 2 * abs(r))
}


panel.cor <- function(x, y, digits=2, prefix="", cex.cor)
{
	usr <- par("usr"); on.exit(par(usr))
	par(usr = c(0, 1, 0, 1))
	r = (cor(x, y,use="pairwise"))
	txt <- format(c(r, 0.123456789), digits=digits)[1]
	txt <- paste(prefix, txt, sep="")
	if(missing(cex.cor)) cex <- 0.5/strwidth(txt)
	text(0.5, 0.5, txt, cex = 2 * abs(r))
}


panel.hist <- function(x, ...)
{
	usr <- par("usr"); on.exit(par(usr))
	par(usr = c(usr[1:2], 0, 1.5) )
	h <- hist(x, plot = FALSE)
	breaks <- h$breaks; nB <- length(breaks)
	y <- h$counts; y <- y/max(y)
	rect(breaks[-nB], 0, breaks[-1], y, col="#009E73", ...)
}


scatterPlotMatrix <- function (x,y,smooth=TRUE,scale=FALSE)
{if (smooth ){
if (scale) {
	pairs(x,diag.panel=panel.hist,upper.panel=panel.cor.scale,lower.panel=panel.smooth)
}
else {pairs(x,diag.panel=panel.hist,upper.panel=panel.cor,lower.panel=panel.smooth)
} 
}
else
{ if (scale) {pairs(x,diag.panel=panel.hist,upper.panel=panel.cor.scale)
} else {pairs(x,diag.panel=panel.hist,upper.panel=panel.cor) }
}
}

scatterPlotMatrix(mtcars)
```   

Plot 1: Scatterplot matrix with correlation and frequency plot.

```{r boxplot, echo=FALSE, fig.width=11, fig.height=5}
library(ggplot2)
p <- ggplot(mtcars, aes(am, mpg))
p + geom_boxplot(aes(fill=am))
```

Plot 2: Boxplot of <code>miles per gallon</code> grouped by <code>transmission</code>.

```{r, echo=FALSE, fig.width=10, fig.height=7}
par(mfrow = c(2,2))
plot(mod3)
```

Plot 3: Residual plot for the best model <code>mod3</code>